---
title: "Reproducible Research Project 1"
author: "Emma Richard"
date: "12/3/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the data

```{r load data, echo=TRUE}
#Load the data 
unzip(zipfile="activity.zip")
activity_data <- read.csv("activity.csv", header = TRUE)
```

## What is mean total number of steps taken per day?


```{r steps, echo=TRUE}
# Calculate the total number of steps taken per day
steps.date <- aggregate(steps~date,activity_data,sum)
# Make a histogram of the total number of steps taken each day
hist(steps.date$steps, xlab="Total Steps per Day", ylab="Frequency [Days]")

#Calculate and report the mean and median of the total number of steps taken per day
meansteps <- mean(steps.date$steps, na.rm=TRUE)
meansteps
mediansteps <- median(steps.date$steps, na.rm=TRUE)
mediansteps

```

## What is the average daily activity pattern?
```{r activitypattern, echo=TRUE}
# Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
steps.interval <- aggregate(steps ~ interval, data=activity_data, FUN=mean)
plot(steps.interval, type="l",xlab="Interval (per 5 min)", ylab="Average Daily Activity Pattern")

# Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
steps.interval$interval[which.max(steps.interval$steps)]
```

## Impute missing values
```{r missing values, echo=TRUE}
# report the total number of missing values in the data
sum(is.na(activity_data))
# fill in missing values - using the mean per 5 minute interval - and create a new dataset that has all missing values filled in
activity_data.new = merge(activity_data, steps.interval, by="interval")
activity_data.new$steps.x[is.na(activity_data.new$steps.x)] = activity_data.new$steps.y[is.na(activity_data.new$steps.x)]
# make a histogram of the total number of steps taken using the new dataset
activity_data.new <- aggregate(steps.x~interval,activity_data.new,sum)
hist(activity_data.new$steps.x, xlab="Total Steps per Day", ylab="Frequency [Days]")
# calculate the mean and median of the new dataset
meansteps_new <- mean(activity_data.new$steps, na.rm=TRUE)
meansteps_new
mediansteps_new <- median(activity_data.new$steps, na.rm=TRUE)
mediansteps_new
```

## Are there differences in activity level between weekdays and weekends?
```{r weekday/weekend, echo=TRUE}
# Create a new factor variable in the dataset with two levels – “weekday” and “weekend” 
daytype <- function(date) {
    if (weekdays(as.Date(date)) %in% c("Saturday", "Sunday")) {
        "Weekend"
    } else {
        "Weekday"
    }
}
activity_data$daytype <- as.factor(sapply(activity_data$date, daytype))
str(activity_data)
# Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
par(mfrow=c(2,1))
for (type in c("Weekend", "Weekday")) {
    steps.type <- aggregate(steps ~ interval,
                            data=activity_data,
                            subset=activity_data$daytype==type,
                            FUN=mean)
    plot(steps.type, type="l", main=type)
}
